---
title: "Seed Theory"
author: "Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

###  Non-spatial individual-based simulation of population dynamics in Daphnia
### https://cran.r-project.org/web/packages/simecol/vignettes/a-simecol-introduction.pdf

## Clear workspace and set directory
```{r setup}
rm(list=ls())
getwd()
setwd("~/GitHub/seedtheory")
```


## Load packages and functions
```{r}
require("png")
require("dplyr")
require("grid")
require("simecol")
```


## Part 1: main equation, parameters, initial values, time steps & solver

```{r}
# Class definition
setClass("indbasedModel", 
         representation(
           parms  = "list",
           init   = "data.frame"
           ),
         contains = "simObj"
         )

# Main equation and parameters
ibm_daphnia <- new("indbasedModel", 
        main = function(time, init, parms) {
          init <- live(init, parms)
          init <- survive(init, parms)
          init <- hatch(init, parms)
          init
          },
        parms = list(
          
          # parameters of the somatic growth equation
          a1 = 1.167,    # (mm)
          a2 = 0.573,    # (mg L^-1)
          a3 = 1.420,    # (mm)
          a4 = 2.397,    # (d),
          b1 = 1.089e-2, # (d^-1)
          b2 = 0.122,    # ((deg. C)^-1)
          
          # parameters of the clutch size equation
          X_max_slope = 23.83,    # (eggs)
          K_s_slope   = 0.65,     # (mg L^-1)
          beta_min    = -29.28,   # (eggs)
          u_c         = 1,        # (L mg^-1) unit conversion factor
          
          # parameters of the individual-based model
          L_0_Hall    = 0.35,     # (mm) SON (size of neonates) of Hall data
          L_0         = 0.65,     # (mm) SON
          SAM         = 1.50,     # (mm) SAM (size at maturity)
          maxage      = 60,       # (d)
          
          # constant environmental conditions
          temp        = 20,       # (deg C)
          food        = 0.5       # (mg L^-1)
          ),

        # initial conditions        
        init      = data.frame(age = 0, size = 0.65, eggs = 0, eggage = 0),
        
        # time steps
        times     = c(from = 0, to = 60, by = 1),
        
        # solver
        solver    = "myiteration", # or default method: "iteration"
        equations = list()
        )
```

## Part 2: Equations and algorithms

```{r}
equations(ibm_daphnia) <- list(
  newdaphnia = function(SON, n) {
    if (n>0) {
      data.frame(age = rep(0, n), size = SON, eggs = 0, eggage = 0)
      } else {
        NULL
        }
    },
  
  bottrell = function(temp) {
    exp(3.3956 + 0.2193 * log(temp) - 0.3414 * log(temp)^2)
    },
  
  tefi = function(time, temp, food, parms)
    {with(parms, {
      deltaL <- L_0 - L_0_Hall
      k      <- b1 * exp(b2 * temp)
      L_max  <- (a1 * food)/(a2 + food) + a3 - k * a4
      L      <- L_max - (L_max - L_0_Hall) * exp (-k * time) + deltaL
      E      <- (X_max_slope * food)/(K_s_slope + food) * L +
                  beta_min * (1 - exp(-u_c * food))
      as.data.frame(cbind(L, E))
    })},
  
    live = function(inds, parms){
      with(parms,{
        ninds       <- nrow(inds)
        inds$age    <- inds$age + DELTAT
        inds$eggage <- ifelse(inds$size > SAM & inds$eggs > 0,
                              inds$eggage + DELTAT, 0)
        tefi_out    <- tefi(inds$age, temp, food, parms)
        inds$size   <- tefi_out$L
        neweggs     <- round(tefi_out$E)
        inds$eggs   <- ifelse(inds$size > SAM & inds$eggage==0,
                              neweggs, inds$eggs)
        inds
        })},
  
    survive  = function(inds, parms) subset(inds, inds$age < parms$maxage),
  
    hatch = function(inds, parms) {
    newinds <- NULL
    with(parms, {
      edt       <- bottrell(temp)
      have.neo  <- which(inds$eggs > 0 & inds$eggage > edt)
      eggs      <- inds$eggs[have.neo]
      new.neo   <- sum(eggs)
      inds$eggs[have.neo]  <-  inds$eggage[have.neo] <- 0
      newinds <- newdaphnia(L_0, new.neo)
      rbind(inds, newinds)
      })
    }
  )
```

## Part 3: Solver function and plot method

```{r}
## a more appropriate solver (note the observer function
myiteration <- function(y, times = NULL, func = NULL, parms = NULL,
                        animate=FALSE, ...) {
  observer <- function(res) {
    # eggs, size, age, eggage
    number   <- nrow(res)
    meansize <- mean(res$size)
    meanage  <- mean(res$age)
    meaneggs <- mean(res$eggs)
    c(number = number, meansize = meansize, meanage = meanage, meaneggs = meaneggs)

  }
  init              <- y@init
  times             <- fromtoby(y@times)
  func              <- y@main
  parms             <- y@parms
  inputs            <- y@inputs
  equations         <- y@equations
  equations         <- addtoenv(equations)
  environment(func) <- environment()
  parms$DELTAT <- 0
  res <- observer(init)
  out <- res
  for (i in 2:length(times)) {
    time <- times[i]
    parms$DELTAT <- times[i] - times[i-1]
    init <- func(time, init, parms)
    res  <- observer(init)
    out  <- rbind(out, res)
   }
 row.names(out) <- NULL
 out <- cbind(times, out)
 as.data.frame(out)
}

## a plotting function that matches the output structure of the observer
setMethod("plot", c("indbasedModel", "missing"), function(x, y, ...) {
  o <- out(x)
  par(mfrow=c(2, 2))
  plot(o$times, o$meanage,  type="l", xlab="Time", ylab="Mean age (d)")
  plot(o$times, o$meaneggs, type="l", xlab="Time", ylab="Eggs per indiv.")
  plot(o$times, o$number,   type="l", xlab="Time", ylab="Abundance")
  plot(o$times, o$number, type="l", xlab="Time", ylab="Abundance", log="y")
})

## RUN the MODEL ##
solver(ibm_daphnia) <- "myiteration"
ibm_daphnia <- sim(ibm_daphnia)
plot(ibm_daphnia)  
```  

## Mortality scenarios

```{r}
# At the first step we create one clone of the daphnia_ibm-object, assign settings common to
# all scenarios and an initial sample population
Sc0 <- ibm_daphnia # no mortality
times(Sc0) <- c(from = 0, to = 30, by = 0.2)
parms(Sc0)[c("temp", "food", "mort")] <- c(15, 0.4, 0.1)
init(Sc0) <- data.frame(age=rep(10, 50), size = rep(2.5, 50),
 eggs=rep(5, 50), eggage=runif(50, 0, 4))
  
# Then we replace the default survive-function with a more general one which depends on a
# user-specified mortality function fmort:
equations(Sc0)$survive = function(inds, parms) {
 abundance <- nrow(inds)
 rnd <- runif(abundance)
 mort <- fmort(parms$mort, inds$size) * parms$DELTAT
 subset(inds, rnd > mort)
}

# Copies of object Sc0 are created and modified according to the scenario specification. In the
# example below we have two functions with constant mortality and two other functions where
# per capita mortality is higher for the larger or smaller individuals, respectively:

Sc1 <- Sc2 <- Sc3 <- Sc0
equations(Sc0)$fmort <- function(mort, x) 0
equations(Sc1)$fmort <- function(mort, x) mort
equations(Sc2)$fmort <- function(mort, x){
 mort * 2 * rank(-x) / (length(x) + 1)
 }
equations(Sc3)$fmort <- function(mort, x){
 mort * 2 * rank(x) / (length(x) + 1)
 }

# Finally, scenarios simulated, either line by line (Section 4.2) or listwise with lapply:

sc <- lapply(list(Sc0=Sc0, Sc1=Sc1, Sc2=Sc2, Sc3=Sc3), sim)
