---
title: "Seed Theory"
author: "Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

###  Spatial individual-based simulation of population dynamics 
### https://bradduthie.github.io/blog/individual-based-models-in-r/#birth

Some notes here:
Great example of how to build IBMs from scratch. 
Should be useful for exploring population and demographic attributes of seed banks. 
Thoughts on how to do this:
1) Using Poisson distribution for both births and deaths. 
2) Model active and dormant individuals separately
3) This means that growth (i.e. gains) will be different, where dormant cells will grow from initiation
4) Need to recover age information! (which is done in the Daphnia example)
5) Death functions for active and dormant will be similar, but with different mean values
6) Explore how different dormant death rates affect dynamics; specifically, constant vs. non-constant
7) In non-constant, I suppose probability of death will be a function of time since entering dormancy
8) Also explore how resuscitation is a function of dormancy 


## Clear workspace and set directory
```{r setup}
rm(list=ls())
getwd()
setwd("~/GitHub/seedtheory")
```


## Load packages and functions
```{r}
require("png")
require("dplyr")
require("grid")
require("simecol")
```

## Create an array of individuals
```{r}
inds <- array(data = 0, dim = c(5, 3)) # x = individuals, y = characteristics
colnames(inds) <- c("characteristic_1", "characteristic_2", "characteristic_3")
rownames(inds) <- c("ind_1", "ind_2", "ind_3", "ind_4", "ind_5")
colnames(inds) <- c("body_mass", "x_loc", "y_loc")
```

## Initial values for individual characteristics
```{r}
inds[, 1] <- rnorm(n = dim(inds)[1], mean = 23, sd = 3) # characteristic 1 = body mass
inds[, 2] <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE); # location of individuals on x-axis
inds[, 3] <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE); # location of individuals on x-axis

# Now plot locations of individuals
plot(x = inds[,2], y = inds[,3], pch = 20, cex = 4, xlim = c(1, 8),
     ylim = c(1, 8), xlab = "x location", mar = c(5, 5, 1, 1),
     ylab = "y location", cex.lab = 1.5, cex.axis = 1.5)

```

## Move an individual (inds[1,]) around
```{r}
x_move     <- sample(x = c(-1, 0, 1), size = 1)
y_move     <- sample(x = c(-1, 0, 1), size = 1)
inds[1, 2] <- inds[1, 2] + x_move
inds[1, 3] <- inds[1, 3] + y_move
```  

## Now move all individuals around
```{r}
x_move     <- sample(x = c(-1, 0, 1), size = dim(inds)[1], replace = TRUE)
y_move     <- sample(x = c(-1, 0, 1), size = dim(inds)[1], replace = TRUE)
inds[, 2] <- inds[, 2] + x_move
inds[, 3] <- inds[, 3] + y_move
```  

## Now lets pull the movement together 
```{r}
movement <- function(inds, xloc = 2, yloc = 3){
    total_inds   <- dim(inds)[1]; # Get the number of individuals in inds
    move_dists   <- c(-1, 0, 1);  # Define the possible distances to move
    x_move       <- sample(x = move_dists, size = total_inds, replace = TRUE);
    y_move       <- sample(x = move_dists, size = total_inds, replace = TRUE);
    inds[, xloc] <- inds[, xloc] + x_move;
    inds[, yloc] <- inds[, yloc] + y_move;
    return(inds);
}
```

## Now let's simulate movement over time
```{r}
time_steps <- 20 # number of total time steps

ts <- 0; # loop calls movement from from time zero to time_steps
while(ts < time_steps){
    inds <- movement(inds);
    ts   <- ts + 1; 
}
```

## This movmement code allows individuals near edges to wrap around
```{r}
movement <- function(inds, xloc = 2, yloc = 3, xmax = 8, ymax = 8){
    total_inds   <- dim(inds)[1]; # Get the number of individuals in inds
    move_dists   <- c(-1, 0, 1);  # Define the possible distances to move
    x_move       <- sample(x = move_dists, size = total_inds, replace = TRUE);
    y_move       <- sample(x = move_dists, size = total_inds, replace = TRUE);
    inds[, xloc] <- inds[, xloc] + x_move;
    inds[, yloc] <- inds[, yloc] + y_move;
    # =========   The reflecting boundary is added below
    for(i in 1:total_inds){               # For each individual i in the array
        if(inds[i, xloc] > xmax){         # If it moved passed the maximum xloc
            inds[i, xloc] <- xmax - 1;    # Then move it back toward the centre
        }
        if(inds[i, xloc] < 1){            # If it moved below 1 on xloc
            inds[i, xloc] <- 2;           # Move it toward the centre (2)
        }
        if(inds[i, yloc] > ymax){         # If it moved passed the maximum yloc
            inds[i, yloc] <- ymax - 1;    # Then move it back toward the centre
        }
        if(inds[i, yloc] < 1){            # If it moved below 1 on yloc
            inds[i, yloc] <- 2;           # Then move it toward the centre (2)
        }
    } 
    # =========  Now all individuals should stay on the landscape
    return(inds);
}
```

## Let's reinitialize the population
```{r}
inds           <- array(data = 0, dim = c(5, 3))
colnames(inds) <- c("body_mass", "x_loc", "y_loc");
rownames(inds) <- c("ind_1", "ind_2", "ind_3", "ind_4", "ind_5")
inds[,1]       <- rnorm(n = dim(inds)[1], mean = 23, sd = 3)
inds[,2]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)
inds[,3]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)
```

## Now let's simulate movement over 20 times steps
```{r}
ts         <- 0
time_steps <- 20
while(ts < time_steps){
    inds <- movement(inds)
    ts   <- ts + 1
}
print(inds)
```

## To track the history of movement, we make some slight modifications
```{r}
ts         <- 0
time_steps <- 20
inds_hist <- NULL # Creates an empty list
while(ts < time_steps){
    inds <- movement(inds)
    ts   <- ts + 1
    inds_hist[[ts]] <- inds; # Add to list
}
#print(inds)
#print(inds_hist) # prints all time points as a list
#print(inds_hist[[18]]) # prints characteristics from timepoint 18
```

## To track the history of movement for an individual
```{r}
ind1_locs <- array(data = NA, dim = c(20, 3)) # creates an empty array
for(i in 1:20){
    ind1_locs[i, 1] <- i                     # Save the time step
    ind1_locs[i, 2] <- inds_hist[[i]][1, 2] # xloc for the time step individual 1
    ind1_locs[i, 3] <- inds_hist[[i]][1, 3] # yloc for the time step individual 1
}
colnames(ind1_locs) <- c("time_step", "x_loc", "y_loc")
# print(ind1_locs)
```

## Simulate individual births: reinitilize array
```{r}
inds           <- array(data = 0, dim = c(5, 4))
colnames(inds) <- c("body_mass", "x_loc", "y_loc", "repr")
rownames(inds) <- c("ind_1", "ind_2", "ind_3", "ind_4", "ind_5")
inds[,1]       <- rnorm(n = dim(inds)[1], mean = 23, sd = 3)
inds[,2]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)
inds[,3]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)
```

## Births function
```{r}
birth <- function(inds, lambda = 0.5, repr_col = 4){
    total_inds       <- dim(inds)[1] # Get the number of individuals in inds
    ind_cols         <- dim(inds)[2] # Total inds columns
    inds[, repr_col] <- rpois(n = total_inds, lambda = lambda)
    total_off        <- sum(inds[, repr_col])
    
    # ---- We now have the total number of new offspring; now add to inds
    new_inds     <- array(data = 0, dim = c(total_off, ind_cols))
    new_inds[,1] <- rnorm(n = dim(new_inds)[1], mean = 23, sd = 3)
    new_inds[,2] <- sample(x = 1:8, size = dim(new_inds)[1], replace = TRUE)
    new_inds[,3] <- sample(x = 1:8, size = dim(new_inds)[1], replace = TRUE)
    
    # ---- Our new offspring can now be attached in the inds array
    inds <- rbind(inds, new_inds)
    return(inds)
}
```

## Now simulate birth following movement over time
```{r}
ts         <- 0
time_steps <- 10
inds_hist  <- NULL
while(ts < time_steps){
    inds            <- movement(inds)
    inds            <- birth(inds)
    ts              <- ts + 1
    inds_hist[[ts]] <- inds
}

ind_abund <- array(data = NA, dim = c(10, 2))
for(i in 1:10){
    ind_abund[i, 1] <- i                      # Save the time step
    ind_abund[i, 2] <- dim(inds_hist[[i]])[1] # rows in inds_hist[[i]]
}
colnames(ind_abund) <- c("time_step", "abundance")
print(ind_abund)
```

## Simulate individual death: reinitialize array
```{r}
inds           <- array(data = 0, dim = c(5, 5))
colnames(inds) <- c("body_mass", "x_loc", "y_loc", "repr", "death")
rownames(inds) <- c("ind_1", "ind_2", "ind_3", "ind_4", "ind_5")
inds[,1]       <- rnorm(n = dim(inds)[1], mean = 23, sd = 3)
inds[,2]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)
inds[,3]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)
```


## Death function
```{r}

### would do something more akin to births above in a non-spatial simulation
### read description in website

death <- function(inds, xlen = 8, ylen = 8, dcol = 5, xcol = 2, ycol = 3){
    for(xdim in 1:xlen){      # For each row `xdim` of the landscape...
        for(ydim in 1:ylen){  # For each col `ydim` of the landscape...
            # Get the total number of individuals on the landscape cell
            on_cell <- sum( inds[, xcol] == xdim & inds[, ycol]  == ydim)
            # Only do something if on_cell is more than one
            if(on_cell > 1){
                # Get all of the occupants (by row number) on the cell
                occupants <- which(inds[, xcol] == xdim & inds[, ycol] == ydim)
                # Sample all but one random occupant to die
                rand_occ  <- sample(x = occupants, size = on_cell - 1)
                # Then add their death to the last column of inds
                inds[rand_occ, dcol] <- 1
            }
        }
    }
    return(inds)
}
```

## Implement death
```{r}
inds <- death(inds = inds)
print(inds)
```

## Recover individuals that survived
```{r}
inds <- inds[inds[, 5] == 0,]
print(inds)
```

## Now, we'll integrate birth and death function with movement
```{r}
# ----- Initialize individuals
inds           <- array(data = 0, dim = c(5, 5))
colnames(inds) <- c("body_mass", "x_loc", "y_loc", "repr", "death")
rownames(inds) <- c("ind_1", "ind_2", "ind_3", "ind_4", "ind_5")
inds[,1]       <- rnorm(n = dim(inds)[1], mean = 23, sd = 3)
inds[,2]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)
inds[,3]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)

# ---- Start the simulation as before
ts         <- 0
time_steps <- 20
inds_hist  <- NULL
while(ts < time_steps){
    inds            <- movement(inds)
    inds            <- birth(inds)
    inds            <- death(inds)
    inds            <- inds[inds[, 5] == 0,] # Retain living
    ts              <- ts + 1 
    inds_hist[[ts]] <- inds
}
```

## Summarize population abundance over time with birth and death functions
```{r}
ind_abund <- array(data = NA, dim = c(20, 2))
for(i in 1:20){
    ind_abund[i, 1] <- i                      # Save the time step
    ind_abund[i, 2] <- dim(inds_hist[[i]])[1] # rows in inds_hist[[i]]
}
colnames(ind_abund) <- c("time_step", "abundance")
print(ind_abund)
```


## Now let's add predators
```{r}
pred           <- array(data = 0, dim = c(5, 5))
colnames(pred) <- c("body_mass", "x_loc", "y_loc", "repr", "death")
rownames(pred) <- c("pred_1", "pred_2", "pred_3", "pred_4", "pred_5")
pred[,1]       <- rnorm(n = dim(pred)[1], mean = 23, sd = 3)
pred[,2]       <- sample(x = 1:8, size = dim(pred)[1], replace = TRUE)
pred[,3]       <- sample(x = 1:8, size = dim(pred)[1], replace = TRUE)
```

## Predator movement...same as prey
```{r}
pred <- movement(pred)
```

## Predation function
```{r}
# there are some details here that may not be relevant to our work

predation <- function(pred, inds, xcol = 2, ycol = 3, rcol = 4, dcol = 5){
    predators   <- dim(pred)[1] # Predator number
    pred[, dcol] <- 1           # Assume dead until proven otherwise
    pred[, rcol] <- 0           # Assume not reproducing until proven otherwise
    for(p in 1:predators){       # For each predator (p) in the array
        xloc   <- pred[p, xcol] # Get the x and y locations
        yloc   <- pred[p, ycol]
        N_prey <- sum( inds[, xcol] == xloc & inds[, ycol] == yloc)
        
        # ----- Let's take care of the predator first below
        if(N_prey > 0){
            pred[p, dcol] <- 0  # The predator lives
        }
        if(N_prey > 1){
            pred[p, rcol] <- 1  # The predator reproduces
        }
        
        # ----- Now let's take care of the prey
        if(N_prey > 0){ # If there are some prey, find them
            prey <- which( inds[, xcol] == xloc & inds[, ycol] == yloc)
            if(N_prey > 2){ # But if there are more than 2, just eat 2
                prey <- sample(x = prey, size = 2, replace = FALSE)
            }
            inds[prey, dcol] <- 1 # Record the prey as dead
        }
    } # We now know which inds died, and which prey died & reproduced
    
    # ---- Below removes predators that have died
    pred         <- pred[pred[,dcol] == 0,] # Only survivors now left
    
    # ----- Below adds new predators based on the reproduction above    
    pred_off     <- sum(pred[, rcol]);
    new_pred     <- array(data = 0, dim = c(pred_off, dim(pred)[2]))
    new_pred[,1] <- rnorm(n = dim(new_pred)[1], mean = 23, sd = 3)
    new_pred[,2] <- sample(x = 1:8, size = dim(new_pred)[1], replace = TRUE)
    new_pred[,3] <- sample(x = 1:8, size = dim(new_pred)[1], replace = TRUE)
    pred         <- rbind(pred, new_pred)
    
    # ----- Now let's remove the prey that were eaten
    inds         <- inds[inds[,dcol] == 0,] # Only living prey left
    # Now need to return *both* the predator and prey arrays
    pred_prey <- list(pred = pred, inds = inds)
    return(pred_prey)
}
```

## New functions
```{r}
pred_prey <- predation(inds = inds, pred = pred)
inds      <- pred_prey$inds
pred      <- pred_prey$pred
```

## Now let's put it all togeteher: movement, prey, and predation 
```{r}
# ----- Initialize individuals (prey); now n = 40 individuals
inds           <- array(data = 0, dim = c(40, 5))
inds[,1]       <- rnorm(n = dim(inds)[1], mean = 23, sd = 3)
inds[,2]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)
inds[,3]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)

# ----- Initialize individuals (predator)' npw n = 20 individuals
pred           <- array(data = 0, dim = c(20, 5))
pred[,1]       <- rnorm(n = dim(pred)[1], mean = 23, sd = 3)
pred[,2]       <- sample(x = 1:8, size = dim(pred)[1], replace = TRUE)
pred[,3]       <- sample(x = 1:8, size = dim(pred)[1], replace = TRUE)

# ---- Start the simulation as before
ts         <- 0
time_steps <- 40
inds_hist  <- NULL
pred_hist  <- NULL
while(ts < time_steps){
    pred            <- movement(pred)
    inds            <- movement(inds) # Note: I increased prey birth rate
    inds            <- birth(inds, lambda = 1.5)
    pred_prey_res   <- predation(pred = pred, inds = inds)
    pred            <- pred_prey_res$pred
    inds            <- pred_prey_res$inds
    inds            <- death(inds)
    inds            <- inds[inds[, 5] == 0,] # Retain living
    ts              <- ts + 1 
    inds_hist[[ts]] <- inds
    pred_hist[[ts]] <- pred
}
```

## From above, calcuate prey and predator abundances
```{r}
ind_abund <- array(data = NA, dim = c(40, 3))
for(i in 1:40){
    ind_abund[i, 1] <- i                      # Save the time step
    ind_abund[i, 2] <- dim(inds_hist[[i]])[1] # rows in inds_hist[[i]]
    ind_abund[i, 3] <- dim(pred_hist[[i]])[1] # rows in pred_hist[[i]]
}
colnames(ind_abund) <- c("time_step", "abundance", "predators")
print(ind_abund)
```


## Plot the results
```{r}
par(mar = c(5, 5, 1, 1))
plot(x = ind_abund[,2], type = "l", lwd = 3, ylim = c(0, 80), 
     xlab = "time step", ylab = "Population abundance", cex.axis = 1.5, 
     cex.lab = 1.5)
points(x = ind_abund[,3], type = "l", lwd = 3, col = "red")
legend(x = 0, y = 80, legend = c("Predator", "Prey"), col = c("red", "black"),
       cex = 1.25, lty = c("solid", "solid"), lwd = c(3, 3))
```

